---
title: "Peer Assessment 1"
author: "cpebenito"
date: "Saturday, April 18, 2015"
output: html_document
---

Loading the Assignment dataset into R

```{r}
activity <- read.csv("activity.csv")
str(activity)
summary(activity)

```


**What is mean total number of steps taken per day?**

*1.Calculate the total number of steps taken per day*

```{r calculating the number of steps}
act2 <- aggregate(. ~ date, data = activity, FUN=sum)

```

*2. Make a histogram of the total number of steps taken each day*

```{r, echo=FALSE}

hist(act2$steps, xlab="Steps", main="Histogram of Steps per Day")

```

*3. Calculate and report the mean and median of the total number of steps taken per day*

```{r mean and median}

mean(act2$steps)
median(act2$steps)

```

**What is the average daily activity pattern?**

```{r}

act3 <- aggregate(. ~ interval, data = activity, FUN=mean)

```

*Series Plot showing the avergae daily activity pattern*

```{r, echo=FALSE}
plot(act3$interval, act3$steps, type="l", xlab="Interval", ylab="Average Steps", main="Time Series of Steps per Interval")

```

*Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*

```{r whichmax}

act4 <- subset(act3, select= -date)
act4[which.max(act4$steps),]
  
```


**Imputing missing values**

```{r how many NAs?}

sum(is.na(activity))


```

```{r results='hide', message=FALSE, warning=FALSE}
if (!require(mice)) {
  install.packages("mice", repos="http://cran.rstudio.com/bin/windows/contrib/3.2/mice_2.22.zip") 
  require(mice)
  library("Rcpp")
  library("lattice")
  library("mice")
}

```


```{r filling missing values, echo=FALSE, results='hide'}
fillup <- activity[c("steps", "date", "interval")]
summary(fillup)
set.seed(200)
imputed <- complete(mice(fillup))
summary(imputed)
```

*Make a histogram of the total number of steps taken each day*

```{r}
impute2 <- aggregate(. ~ date, data = imputed, FUN=sum)

hist(impute2$steps, xlab="Steps", main="Histogram of Steps per Day")

```

*Calculate and report the mean and median total number of steps taken per day, which include imputed values*

```{r}

mean(impute2$steps)
median(impute2$steps)

```

**What is the impact of imputing missing data on the estimates of the total daily number of steps?**

```{r histograms}

if (!require(ggplot2)) {
    install.packages("ggplot2", repos="http://cran.us.r-project.org")
    require(ggplot2)
}
library(ggplot2)
set.seed(800)

act2$source <- 'withNAs'
impute2$source <- 'noNAs'
dat <- rbind(act2, impute2)

```
  
  In the histogram shown below, the blue lines represent the mean (solid) and median (dotted) of the dataset withNAs and the red lines represent the dataset noNAs. There was a slight increase on the values when the NAs were imputed using the MICE(Multivariate Imputation by Chained Equations) Package in R.
 
 
```{r histogram showing difference between imputed value and original value}
 
 ggplot(dat, aes(steps, fill=source)) + geom_histogram( alpha=0.5 ) +
  geom_vline(aes(xintercept=mean(steps)), data=act2, color="blue") +
  geom_vline(aes(xintercept=mean(steps)), data=impute2, color="red") +
  geom_vline(aes(xintercept=median(steps)), data=act2, color="blue", 
             linetype="dotted", size=1) +
  geom_vline(aes(xintercept=median(steps)), data=impute2, color="red", 
             linetype="dotted", size=1)
```

  
```{r results comparison and rounding of values, echo = FALSE}
  
mean_withNA <- round(mean(act2$steps), digits = 0)
median_withNA <- round(median(act2$steps), digits = 0)
mean_noNA <- round(mean(impute2$steps), digits = 0)
median_noNA <- round(median(impute2$steps), digits = 0)

```

Again for purposes of comparison mean(with NAs) = `r mean_withNA` while mean(no NAs) = `r mean_noNA`. On the other hand, median(with NAs) = `r median_withNA` while median(no NAs) = `r median_noNA`



**Are there differences in activity patterns between weekdays and weekends?**

*The time series plot below shows the activity patterns between a weekday and weekend activity.*

```{r, echo=FALSE}
## I'm sure there's a better way of classifying weekday and weekend, but this is the best I could do for now with the deadline ##

imputed$date2 <- as.POSIXct(imputed$date, tz="")
imputed$day <- weekdays(imputed$date2)

imputed$dayno<- factor(imputed$day,
                              levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), labels = c(1,2,3,4,5,6,7))
                         
imputed$wcat <- as.integer(imputed$dayno)
  
imputed$weekcat <- ifelse(imputed$wcat <= 4, "weekday", "weekend")


g <- ggplot(imputed, aes(interval, steps))
g + geom_line() + facet_wrap( ~ weekcat, ncol = 1) + labs(title = "Activity Pattern")


```



